Spatiotemporal model of `r paste(params$species)` occupancy

### 2020-11-25 analysis/sand-lance/biomass-density-model.Rmd
refine models and effect plots

---
title: "Spatiotemporal model of `r paste(params$species)` biomass"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    covs: "-mud-depth-temp"
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

@@

```{r params}
species <- params$species
region <- params$region
covs <- params$covs
knots <- params$knots
REML <- params$REML
AR1 <- params$AR1

paste("region =", region)
paste("climate covariate =", clim_cov)
paste("model label =", covs)
paste("knots =", knots)
paste("REML =", REML)
paste("AR1 =", AR1)
```

@@

d <- left_join(data, bath) %>% mutate(
  depth_sam = depth,# rename sensor depth
  depth = coalesce(depth, depth_bath),# use bath depth if sensor depth missing
  raw_depth = depth, # create 'depth' var that's actually log depth for scaling
  log_depth = log(depth), depth = log(raw_depth)) %>% 
  rename(temp = temperature_c, do = do_mlpl)

@@

```{r eval= FALSE}
# depth diff = difference between bath and sample depths
# depth range is difference between start and end sample depths
ggplot(data, aes(depth_sam, depth_bath, colour = depth_range)) + 
  geom_point(alpha=0.75) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1) +
  geom_abline(intercept = 0, slope = 1)

@@

@@

    total_biomass <- sdmTMB::sdmTMB(
      data = data, mod.formula,
      # time_varying = ~ 0 + depth_scaled + depth_scaled2, 
      time = "year_pair", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = params$AR1,
      reml = params$REML,
      include_spatial = params$fixed_spatial
    )

@@

```{r}
g2 <- ggplot(pred, 
      aes((est), residuals, colour = density)) +
    geom_point(alpha = 0.2) + # scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g2
```

@@

### Effect plots
```{r}
if(params$update_model_check){

nd_depth <- data.frame(
  depth_scaled = seq(min(data$depth_scaled), 
    max(data$depth_scaled), length.out = 30), 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = 0,
  roms_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_depth <- ggplot(p_depth, aes(depth, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max(exp(p_depth$est))*20), xlim = c(0, 150)) +
  geom_line() + geom_ribbon(alpha = 0.4))

@@

```{r}
if(clim_cov == "roms"){
nd_roms <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = 0,
  muddy_scaled = 0,
  roms_scaled = seq(min(data$roms_scaled), 
    max(data$roms_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_roms <- predict(m, newdata = nd_roms, se_fit = TRUE, re_form = NA)
p_roms$roms <- (p_roms$roms_scaled*data$roms_sd[1])+ data$roms_mean[1]

@@

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

(g_temp <- ggplot(p_temp, aes(temp, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0, max(exp(p_temp$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

@@

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_sand <- ggplot(p_sand, aes(sandy, exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0,max(exp(p_sand$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

# ggplot(p_sand, aes(sandy, exp(est),
#   ymin = exp(est - 1.96 * est_se), 
#   ymax = exp(est + 1.96 * est_se))) +
#   coord_cartesian(ylim = c(0, max(exp(p_sand$est))*2), xlim = c(0,1)) +
#   geom_line() + geom_ribbon(alpha = 0.4)

@@

```{r}
if(covs %in% c("-mud-depth-roms", "-mud-depth-temp")){
nd_mud <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  muddy_scaled = seq(min(data$muddy_scaled), 
    max(data$muddy_scaled), length.out = 50),
  sandy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_mud <- ggplot(p_mud, aes(muddy, exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0,max(exp(p_mud$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

# ggplot(p_mud, aes(muddy, exp(est),
#   ymin = exp(est - 1.96 * est_se), 
#   ymax = exp(est + 1.96 * est_se))) +
#   coord_cartesian(ylim = c(0, max(exp(p_mud$est))*2), xlim = c(0,1)) +
#   geom_line() + geom_ribbon(alpha = 0.4)

>>>>

---
title: "Spatiotemporal model of `r paste(params$species)` biomass"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    covs: "-sand-depth-temp"
    # covs: "-mud-depth-temp"
    # covs: "-mud-sand-depth-temp"
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

if(params$covs == "-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, by = shallow, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}
if(params$covs == "-mud-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

@@

```{r params}
species <- params$species
region <- params$region
covs <- params$covs
knots <- params$knots
# REML <- params$REML
# AR1 <- params$AR1

paste("region =", params$region)
paste("climate covariate =", clim_cov)
paste("model label =", params$covs)
paste("knots =", params$knots)
paste("REML =", params$REML)
paste("AR1 =", params$AR1)
paste("fixed spatial =", params$fixed_spatial)
paste("formula =", mod.formula[3])
```

@@

d <- left_join(data, bath) %>% mutate(
  depth_sam = depth,# rename sensor depth
  depth = coalesce(depth, depth_bath),# use bath depth if sensor depth missing
  raw_depth = depth, # create 'depth' var that's actually log depth for scaling
  log_depth = log(depth), depth = log(raw_depth),
  shallow = if_else(raw_depth < 100, 1, 0)
  ) %>% 
  rename(temp = temperature_c, do = do_mlpl)

@@

```{r eval = FALSE}
# depth diff = difference between bath and sample depths
# depth range is difference between start and end sample depths
ggplot(data, aes(depth_sam, depth_bath, colour = depth_range)) + 
  geom_point(alpha=0.75) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1) +
  geom_abline(intercept = 0, slope = 1)

@@

Check for substrate correlations -- sandy somewhat correlated with depth
```{r eval= FALSE}
ggplot(data, aes(raw_depth, sandy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

ggplot(data, aes(raw_depth, sandy-muddy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

ggplot(data, aes(raw_depth, muddy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

ggplot(filter(data, raw_depth < 100), aes(depth, sandy, colour = depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1) 
```

@@

    total_biomass <- sdmTMB::sdmTMB(
      data = data, mod.formula,
      time = "year_pair", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = params$AR1,
      reml = params$REML,
      include_spatial = params$fixed_spatial
    )

@@

```{r}
g2 <- ggplot(pred, 
      aes(exp(est), residuals, colour = density)) +
    geom_point(alpha = 0.2) + 
    scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g2
```

@@

### Effect plots
```{r}
threshold <- quantile(data$density, 0.999)
if(params$update_model_check){

nd_depth <- data.frame(
  depth_scaled = seq(min(data$depth_scaled), 
    max(data$depth_scaled), length.out = 30), 
  shallow = 1,
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = 0,
  roms_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

max_CI <- max(exp(p_depth$est + 1.96 * p_depth$est_se))
max_Y <- if_else(threshold > max_CI, max_CI, threshold)

(g_depth <- ggplot(p_depth, aes(depth, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max_Y), xlim = c(0, 150)) +
  geom_line() + geom_ribbon(alpha = 0.4))

@@

```{r}
if(clim_cov == "roms"){
nd_roms <- data.frame(
  depth_scaled = min(data$depth_scaled), # use min depth for estimated prob at most relevant depth
  shallow = 1,
  sandy_scaled = 0,
  muddy_scaled = 0,
  roms_scaled = seq(min(data$roms_scaled), 
    max(data$roms_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_roms <- predict(m, newdata = nd_roms, se_fit = TRUE, re_form = NA)
p_roms$roms <- (p_roms$roms_scaled*data$roms_sd[1])+ data$roms_mean[1]

max_CI <- max(exp(p_roms$est + 1.96 * p_roms$est_se))
max_Y <- if_else(threshold > max_CI, max_CI, threshold)

@@

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  shallow = 1,
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

max_CI <- max(exp(p_temp$est + 1.96 * p_temp$est_se))
max_Y <- if_else(threshold > max_CI, max_CI, threshold)

(g_temp <- ggplot(p_temp, aes(temp, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max_Y)) +
  geom_line() + geom_ribbon(alpha = 0.4))

@@

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp", "-mud-sand-depth-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  shallow = 1,
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = min(data$muddy_scaled),
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_sand <- ggplot(p_sand, aes(sandy, exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max_Y)) +
  geom_line() + geom_ribbon(alpha = 0.4))

@@

```{r}
if(covs %in% c("-mud-depth-roms", "-mud-depth-temp","-mud-sand-depth-temp")){
nd_mud <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  shallow = 1,
  muddy_scaled = seq(min(data$muddy_scaled), 
    max(data$muddy_scaled), length.out = 50),
  sandy_scaled = max(data$sandy_scaled),
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

max_CI <- max(exp(p_mud$est + 1.96 * p_mud$est_se))
max_Y <- if_else(threshold > max_CI, max_CI, threshold)

(g_mud <- ggplot(p_mud, aes(muddy, exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max_Y)) +
  geom_line() + geom_ribbon(alpha = 0.4))

### 2020-11-25 analysis/sand-lance/occupancy-model.Rmd
refine models and effect plots

---
title: "Spatiotemporal model of `r paste(params$species)` occupancy"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    # covs: "-mud-depth-temp"
    # covs: "-sand-mud-depth-temp"
    covs: "-sand-by-depth-mud-temp"
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

if(params$covs == "-sand-by-depth-mud-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  te(sandy_scaled, depth_scaled, k = 5) + 
  s(muddy_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

@@

```{r params}
species <- params$species
region <- params$region
covs <- params$covs
knots <- params$knots
REML <- params$REML
AR1 <- params$AR1

paste("region =", region)
paste("climate covariate =", clim_cov)
paste("model label =", covs)
paste("knots =", knots)
paste("REML =", REML)
paste("AR1 =", AR1)
paste("formula =", mod.formula)
```

@@

```{r}
if(clim_cov == "roms"){
nd_roms <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = 0,
  muddy_scaled = 0,
  roms_scaled = seq(min(data$roms_scaled), 
    max(data$roms_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_roms <- predict(m, newdata = nd_roms, se_fit = TRUE, re_form = NA)
p_roms$roms <- (p_roms$roms_scaled*data$roms_sd[1])+ data$roms_mean[1]

@@

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, # use min depth for estimated error at relevant depths
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

@@

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

p_mean <- predictions  %>% group_by(X, Y) %>%
  mutate(mean_est = mean(logit2prob(est))) %>% 
  filter(year_pair == 2011) %>% 
  mutate(year = "all years") %>%
  plot_facet_map("mean_est",
    raster_limits = c(0, max_raster),
    transform_col = fourth_root_power,
    legend_position = legend_coords
  ) + labs(fill = "prob") +
  ggtitle(paste0("Estimated biomass density of ", species, 
    " \n(max = ", max_prob, ")"))
print(p_mean)
ggsave(paste0("figs/map-", spp, covs, "-mean-distribution.png"), width = 5, height = 5)
```

>>>>

---
title: "Spatiotemporal model of `r paste(params$species)` occupancy"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    covs: "-mud-depth-temp"
    # covs: "-sand-mud-depth-temp"
    # covs: "-sand-by-depth-mud-temp"
    REML: FALSE
    AR1: FALSE # true doesn't seem to work with occupancy models
    fixed_spatial: FALSE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

if(params$covs == "-sand-by-depth-mud-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  te(sandy_scaled, depth_scaled, k = 3) + 
  s(muddy_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

@@

```{r params}
species <- params$species
region <- params$region
covs <- params$covs
knots <- params$knots
# REML <- params$REML
# AR1 <- params$AR1

paste("region =", params$region)
paste("climate covariate =", clim_cov)
paste("model label =", params$covs)
paste("knots =", params$knots)
paste("REML =", params$REML)
paste("AR1 =", params$AR1)
paste("fixed spatial =", params$fixed_spatial)
paste("formula =", mod.formula[3])
```

@@

```{r}
if(clim_cov == "roms"){
nd_roms <- data.frame(
  depth_scaled = min(data$depth_scaled), # use min depth for estimated prob at most relevant depth
  sandy_scaled = 0,
  muddy_scaled = 0,
  roms_scaled = seq(min(data$roms_scaled), 
    max(data$roms_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_roms <- predict(m, newdata = nd_roms, se_fit = TRUE, re_form = NA)
p_roms$roms <- (p_roms$roms_scaled*data$roms_sd[1])+ data$roms_mean[1]

@@

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

@@

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

p_mean <- predictions  %>% group_by(X, Y) %>%
  mutate(mean_est = mean(logit2prob(est))) %>% 
  filter(year_pair == 2011) %>% 
  mutate(year = "all years") %>%
  plot_facet_map("mean_est",
    raster_limits = c(0, max_raster),
    transform_col = fourth_root_power,
    legend_position = legend_coords
  ) + labs(fill = "prob") +
  ggtitle(paste0("Estimated occupancy of ", species, 
    " \n(max = ", max_prob, ")"))
print(p_mean)
ggsave(paste0("figs/map-", spp, covs, "-mean-occupancy.png"), width = 5, height = 5)
```
